# Reproducible Research Peer Assessment 1

## by LRG2000

```{r setoptions, echo = TRUE, message = FALSE}
opts_chunk$set(echo = TRUE)
opts_chunk$set(message = FALSE)
```

First, read in the data:
```{r readData}
data <- read.csv("activity.csv", stringsAsFactors = FALSE)
```

Now we can begin to answer questions about this dataset. The data consist of the number of steps taken summed over 5-minute intervals. The first "question"" asks for a histogram of the total number of steps taken each day, the median and mean number of steps taken each day. 

```{r separate_days, fig.height = 5}
s <- split(data, data$date)
dailyVal <- numeric(length(s))
for (i in 1:length(s)) {dailyVal[i] <- sum(s[[i]]$steps[!is.na(s[[i]]$steps)])}
hbreaks <- c(0, 40, 3000, 6000, 9000, 12000, 15000, 18000, 22000)
plotbreaks <- c(0, 1480, 4500, 7500, 10500, 13500, 16500, 20000)
hist(dailyVal, breaks=hbreaks, freq=TRUE, xaxt="n", xlab="number of steps", ylim=c(0,20))
axis(side=1,at=plotbreaks, labels=as.character(plotbreaks))
abline(v=mean(dailyVal), col="red", lwd=2)
text(mean(dailyVal)-1700, y=20, labels="mean")
abline(v=median(dailyVal), col="blue", lwd=2)
text(median(dailyVal)+2200, y=20, labels="median")

```

We can also look at the average number of steps in each time interval. The time intervals listed below begin at the 24-hour time 
specified.

```{r}
interval_a <- as.character(10000+s[[1]]$interval)
interval <- character(288)
intAve <- numeric(288)
int.df <- data.frame(s)
steps <- NULL
for (i in 1:288){ ## get the interval number to look like a time
      interval[i] <- paste(substr(interval_a[i], 2,3),":", substr(interval_a[i], 4, 5), sep="", collapse="")}
for (i in 1:61) { steps <- rbind(steps, s[[i]]$steps)}
means <- colMeans(steps, na.rm = TRUE)
plot(s[[1]]$interval, means, xaxt = "n", type="l", xlab="interval", ylab="averave # of steps")
axis(side=1, labels = interval[c(1,(1+18*(1:12)))], at=s[[1]]$interval[c(1,(1+22*(1:12)))], lwd.ticks = 1)
```
In our calculation of mean number of steps in an interval, we have avoided missings values. As can be seen from the first histogram, there are eight days with steps measured the entire day. Now we are going to examine how many missing values there are. The interval with the largest average number of steps occurred at `r interval[which.max(means)]`

```{r}
ind1 <- is.na(data$steps); ind2 <- is.na(data$interval[!ind1])
nas <- sum(ind1) + sum(ind2)
```
After checking the numbers of NAs in the individual elements, we found that there are `r nas` NAs in the "steps" element, and none in the other two list elements. We're going replace any NA values in an interval with the mean number of steps in that interval. Then we will calculate the total number of steps taken each day with these values.

```{r}
nsteps <- split(data$steps, data$interval)
for (i in 1:288) {
      ind <- is.na(nsteps[[i]])
      if (sum(ind > 0)) {
            nsteps[[i]][ind] <- means[i]}
}
newDailyVal <- numeric(61)
for (i in 1:61) {
      s[[i]]$steps[is.na(s[[i]]$steps)] <- means[is.na(s[[i]]$steps)]
      newDailyVal[i] <- sum(s[[i]]$steps)}

hist(newDailyVal, breaks=hbreaks, freq=TRUE, xaxt="n", xlab="number of steps", ylim=c(0,28), main = "histogram without NAs")
axis(side=1,at=plotbreaks, labels=as.character(plotbreaks))
abline(v=mean(newDailyVal), col="red", lwd=2)
text(mean(newDailyVal)-1700, y=20, labels="mean")
abline(v=median(newDailyVal), col="blue", lwd=2)
text(median(newDailyVal)+2200, y=20, labels="median")
```
We see that the mean and median are now in the same place, at `r sprintf("%5.0f",mean(newDailyVal))` . The mean before replacing NAs was `r sprintf("%4.0f",mean(dailyVal))`, while the old median was `r sprintf("%5.0f",median(dailyVal))`.

Now we will look for differences in activity on weekends and weekdays.
```{r fig.height=10}
us <- unsplit(s, data$date)
date <- as.Date(us$date)
days <- weekdays(date)
wkends <- grepl("Sunday",days) | grepl("Saturday", days)
dayType <- character(length(us$date))
dayType[wkends] <- "weekend"
dayType[!wkends] <- "weekday"
us <- cbind(us, dayType)
wks <- split(us, dayType)
dayInts <- split(wks[["weekday"]],wks[["weekday"]]$interval)
endInts <- split(wks[["weekend"]],wks[["weekend"]]$interval)

plotme <- data.frame(aveSteps = numeric(576), dayType = character(576), interval = rep(s[[1]]$interval,2), 
                     stringsAsFactors=FALSE)
for (i in 1:288) {
      plotme$aveSteps[i] <- mean(dayInts[[i]]$steps)
      plotme$aveSteps[288+i] <- mean(endInts[[i]]$steps)
      plotme$dayType[i] <- "weekday"
      plotme$dayType[i+288] <- "weekend"
      }

library(ggplot2)
qplot(interval, aveSteps, data=plotme, geom="line") + geom_line() + facet_grid(dayType ~.)
```
We find that on weekends, the interval with the average number of steps occurs at `r interval[which.max(plotme$aveSteps[289:576])]`, while on weekdays the interval occurred at `r interval[which.max(plotme$aveSteps[1:288])]`, the same as the whole-week average.

